Type theory in type theory with single substitutions

### 2024-09-13 types_2024_postproc/p.tex
stuff

The traditional way of defining a programming language formally
\cite{DBLP:books/cu/Ha2016,Pierce:SF2,plfa22.08} is comprised of three
steps: (i) definition of the presyntax (abstract syntax trees, ASTs)
as a formal grammar; (ii) specification of
well-formedness/well-typedness relations saying which ASTs are
well-behaved; (iii) specification of rewriting/conversion relations
explaining how programs are executed. Instead of following this route,
we use the intrinsic approach
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
we define ASTs as an inductive type instead of production rules of a
grammar; (ii) terms are indexed by (object theoretic) types, thus only
allowing ASTs to be built in well-typed ways; (iii) terms are
quotiented by conversion rules, making convertible terms equal. The
advantage of the intrinsic approach is that it provides shorter,
easier to understand, less technical definitions, and there is no
mismatch between syntax and semantics. Type theory is a generalised
algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}), the
notion of model is direct, and the syntax is simply the initial
model. GATs are generalisations of single-sorted algebraic theories
(such as monoids, groups) to multiple sorts where later sorts can be
indexed over previous sorts. An example is categories with a sort of
objects and a sort of morphisms double-indexed over objects. In a
type-theoretic metatheory, the initial model of a GAT is a quotient
inductive-inductive type (QIIT
\cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general family of
inductive types where multiple types are defined mutually, later types
can be indexed over previous ones, and equality constructors are also
allowed, generalising quotients. The elimination principle of the QIIT
corresponds to a structural induction principle and is
equivalent to the statement that the syntax is the initial model
\cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a unique
morphism from the syntax to any other model.

There are several ways to define type theory as a GAT. For example,
Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
or C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. These are providing the
basic substitution calculus part of the theory, and individual type
formers (such as function space or product types) have to be specified
separately. With the exception of B-systems, the listed definitions
all define a parallel substitution calculus, where contexts and
substitutions form a category. Like B-systems, in this paper we define
type theory as a single substitution calculus (SSC). Single
substitutions don't compose, so they do not form a category. Our
definition is simpler than B-systems, but it is closely related to
it. It is a minimalistic presentation where every operation and
equation is well-motivated (see Section \ref{sec:ssc}). In addition to
its pedagogical value due to its simplicity, our definition might
contribute in the future to a possible coherent syntax for type
theory, see Section \ref{sec:conclusion}.

Our theory is in some sense too minimalistic: for example, consider a
term $b$ which depends on a context $\Gamma$ and an extra variable of
type $A$, that is, $b : \Tm\,(\Gamma\ext A)\,B$. We can weaken $b$ so
that we add an extra $A$ variable \emph{before} the last one obtaining
$b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p])$ where $\p$ is
weakening at the end of the context, and $\p^+$ is weakening just
before the last variable in the context. Then, we instantiate the last
$A[\p]$ variable in the context by the previous $A$ obtaining
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$ is De
Bruijn index $0$. We cannot derive $b[\p^+][\langle\q\rangle] = b$ in
an arbitrary model, but this equation holds in the syntax.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold for syntactic terms. In fact, all the
operations and equations of the parallel substitution calculus (CwF)
are admissible in our syntax, moreover the set of SSC-based syntactic
terms are isomorphic to the set of CwF-based terms. In contrast with
B-systems which are equivalent to CwFs, our SSC-models do not form
CwF-based models. Every CwF-based model is a model of our SSC, but not
the other way: there are more SSC-models than CwF-models. The
situation is analogous to the relationship of extensional combinatory
calculus and lambda calculus \cite{DBLP:conf/fscd/AltenkirchKSV23}:
the syntactic terms form isomorphic sets, but there are more
combinatory models than lambda calculus models.

@@

In this paper, we define the SSC and extend it with
$\Pi$ types and a hierarchy of Coquand-universes
\cite{coquandUniverse}. The rules for type formers are the same as for
CwFs, the difference is only in the substitution calculus part. More
generally, every second-order generalised algebraic theory (SOGAT) has
a parallel and a single substition-based first order presentation
\cite{DBLP:conf/fscd/KaposiX24}. Here we prove that the syntaxes of
the parallel and the single substitution calculi are isomorphic for
the following SOGAT (we use implicit arguments and $\cong$ denotes
isomorphism):
\begin{alignat*}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
\end{alignat*}
We conjecture that the methods of in this paper can be used to show
that for any SOGAT, the two first-order presentations give isomorphic
syntaxes.

@@

Most formalisation of type theory use low-level

@@

@@

@@

>>>>

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic. That is, we only consider well-formed,
  well-scoped and well-typed abstract syntax trees, there are no
  meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{alignat*}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
\end{alignat*}
We use implicit arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly,
and $\app$ takes $i$, $A$ and $B$ implicitly. $f : X \cong Y : g$
denotes an isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with
$g\,(f\,x) = x$ and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$
and $\lam$ are second-order functions. We say that $\Pi$ binds a
$\Tm$-variable in its second $\Ty$-argument. Every binder is
represented by a second-order function, and as a result, we don't need
to talk about contexts and substitutions when describing the theory:
we re-use the binding features of the metatheory to describe those of
our object theory (this technique is also called higher-order abstract
syntax \cite{DBLP:conf/lics/Hofmann99} or logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
is no good notion of homomorphism between second-order models, thus
the semantics of a SOGAT is given by a first order generalised
algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}). GATs are
generalisations of single-sorted algebraic theories (such as monoids,
groups) to multiple sorts where later sorts can be indexed over
previous sorts, and the usual results from universal algebra (e.g.\ a
category of models with finite limits and initial object, free and
co-free models) carry over to GATs
\cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
theory of categories where there is a sort for objects and a sort for
morphisms double-indexed over objects.

Kaposi and Xie \cite{DBLP:conf/fscd/KaposiX24} present two different
first order semantics for any SOGAT: one based on parallel
substitutions, and one based on single substitutions.

In the parallel semantics, the GAT always starts with a category with
a terminal object, objects of the category are called contexts,
morphisms are substitutions. Every sort and operation of the SOGAT
gives rise to an operation in the GAT which is indexed by
contexts. The sorts moreoever come with functorial instantiation
(sometimes called substitution) operations, and all the operations are
natural with respect to instantiation. For those sorts which are bound
by an operation, we have context extension, meaning that contexts can
carry variables of that sort (in other words, the sort is given by a
locally representable presheaf). The parallel semantics of the GAT
\begin{alignat*}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set 
\end{alignat*}
is the GAT of categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}). To define the notion
of model of a language (given by a SOGAT) is to take its parallel
substitution semantics, which is a GAT, and that induces a notion of
model. The syntax of the language is the initial model, which exists
for every GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an
intrinsic definition of syntax (only well-typed terms), which is
quotiented by the equations in the SOGAT, which describe conversion
for the language.

The single substitution semantics is an alternative GAT xxx

In a type-theoretic metatheory,
the initial model of a GAT is a quotient inductive-inductive type
(QIIT \cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general
family of inductive types where multiple types are defined mutually,
later types can be indexed over previous ones, and equality
constructors are also allowed, generalising quotients. The elimination
principle of the QIIT corresponds to a structural induction principle
and is equivalent to the statement that the syntax is the initial
model \cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a
unique morphism from the syntax to any other model.

conversion to GAT, algebra, models

The traditional way of defining a programming language formally
\cite{DBLP:books/cu/Ha2016,Pierce:SF2,plfa22.08} is comprised of three
steps: (i) definition of the presyntax (abstract syntax trees, ASTs)
as a formal grammar; (ii) specification of
well-formedness/well-typedness relations saying which ASTs are
well-behaved; (iii) specification of rewriting/conversion relations
explaining how programs are executed. Instead of following this route,
we use the intrinsic approach
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
we define ASTs as an inductive type instead of production rules of a
grammar; (ii) types are part of the definition and terms are indexed
by them, thus only allowing ASTs to be built in well-typed ways; (iii)
terms are quotiented by conversion rules, making convertible terms
equal. Compared to the traditional extrinsic approach, the intrinsic approach provides
shorter, less technical definitions, and there
is no mismatch between syntax and semantics. Type theory is a
generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}), the notion of model is direct,
and the syntax is simply the initial model.

There are several ways to define type theory as a GAT. For example,
Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
or C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.  With the exception of
B-systems, the listed notions all define a parallel substitution
calculus, where contexts and substitutions form a category. Like
B-systems, in this paper we define a single substitution calculus
(SSC). Single substitutions don't compose, so they do not form a
category. Our definition is simpler than B-systems, but it is closely
related to it. It is a minimalistic presentation where every operation
and equation is well-motivated (see Section \ref{sec:ssc}). In
addition to its pedagogical value due to its simplicity, in the future
it might give rise to new models or contribute to a possible coherent
syntax for type theory, see Section \ref{sec:conclusion}.

CwFs, C-systems, B-systems and our SSC only provide the basic
substitution calculus part of the theory, and individual type formers
(such as function space or product types) have to be specified
separately. There is a systematic way to do this: non-substructural
programming languages can be specified as second-order generalised
algebraic theories (SOGATs \cite{DBLP:conf/fscd/KaposiX24}), which are
GATs allowing second-order operations. For example, type theory with
$\Pi$ and a family of

Kaposi and Xie \cite{DBLP:conf/fscd/KaposiX24} showed that any SOGAT
has two kinds of GAT semantics, one based on a parallel substitution
calculus, and another based on a single substitution calculus. The
parallel substitution GAT semantics of the SOGAT containing only the
sorts of types and terms ($\Ty : \Set, \Tm : \Ty\ra\Set$) is simply
the GAT of CwFs. The single substitution semantics of the same SOGAT
is the SSC in this paper.

SOGAT stuff

Our SSC is in some sense too minimalistic: for example, consider a
term $b$ which depends on a context $\Gamma$ and an extra variable of
type $A$, that is, $b : \Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so
that we add an extra $A$ variable \emph{before} the last one obtaining
$
b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p])
$
where $\p$ is
weakening at the end of the context, and $\p^+$ is weakening just
before the last variable in the context. Then, we instantiate the last
$A[\p]$ variable in the context by the previous $A$ obtaining
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$ denotes the last
variable in the context (a.k.a\ De Bruijn index $0$).
We cannot derive $b[\p^+][\langle\q\rangle] = b$ in
an arbitrary model, but this equation holds in the syntax.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In fact, we can show
that the syntax xxx

In fact, all the
operations and equations of the parallel substitution calculus (CwF)
are admissible in the syntax of SSC, moreover the set of syntactic SSC
terms are isomorphic to the set of CwF terms. In contrast with
B-systems which are equivalent to CwFs, our SSCs do not form
CwFs. Every CwF-based model is a model of our SSC, but not the other
way: there are more SSC-models than CwF-models. The situation is
analogous to the relationship of extensional combinatory calculus and
lambda calculus \cite{DBLP:conf/fscd/AltenkirchKSV23}: the syntactic
terms form isomorphic sets, but there are more combinatory models than
lambda calculus models.

@@

In this paper, we define the SSC and extend it with
$\Pi$ types and a hierarchy of Coquand-universes
\cite{coquandUniverse}. The rules for type formers are the same as for
CwFs, the difference is only in the substitution calculus part. More
generally, every second-order generalised algebraic theory (SOGAT) has
a parallel and a single substition-based first order presentation
\cite{DBLP:conf/fscd/KaposiX24}. Here we prove that the syntaxes of
the parallel and the single substitution calculi are isomorphic for
the following SOGAT

We conjecture that the methods of in this paper can be used to show
that for any SOGAT, the two first-order presentations give isomorphic
syntaxes.

@@

In the 1990s single substitution calculi for type theory were popular,
but usually in the extrinsic setting, see e.g.\ the thesis of
Altenkirch \cite{alti:phd93}. B-systems were introduced by Voevodsky
\cite{voevodsky2014bsystems} as an algebraic way of describing type
theories close to the notations using typing judgements. Using our
notations B-systems can be described as follows. B-frame.

Ehrhard

we use Sub, but we could have used telescopes

\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}

contextual sCwF and contextual SSC with fct types

Formalisations of type theory use low-level

@@

Initiality conjecture.

@@

\section{Minimisation}

\section{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}

@@

Why are we still looking at alternative ways of defining type theory?
Closer to the syntax/implementation, easier to understand/implement,
fewer equations, hence easier to define models, coherent syntax?
